Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-23 paper/paper.tex
Typo
remove eta from section 5

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

@@

\section{Toward Decidability of Type Conversion and $\eta$ laws}
\label{sec:extension}
In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended show the existence of
$\beta\eta$ normal forms for (open and closed) well-typed terms. We
first extend parallel reduction to include $\eta$ reduction for
functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}

\begin{figure}[h]
  \[
    \begin{array}{lcl}
      % \beta\text{-}\mathit{neutral\ terms}\\
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      % \beta\text{-}\mathit{normal\ terms} \\
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}

The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the neutral terms
and normal forms with respect to $\beta$-reduction, but can still take
$\eta$ reduction steps. We sometimes use the judgment forms $[[ne a]]$
and $[[nf a]]$ to indicate that there exists $[[e]]$ or $[[f]]$ such
that $[[a = e]]$ or $[[a = f]]$.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  natural numbers as term variables. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}
\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}
\begin{proof}
  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
  existence of $\beta\eta$-normal forms for well-typed terms since
  $\eta$-reduction preserves $\beta$-normal form (implied by
  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
  the term (evident from \rref{P-AbsEta}).
\end{proof}

Due to the non-deterministic nature of parallel reduction, we need to
take a few more steps to convert the existence of $\beta\eta$-normal
form into a decision procedure for type conversion. More specifically,
we can show that a deterministic evaluation strategy such as
leftmost-outermost reduction can always find the $\beta\eta$-normal
form if there exists one. Then the termination of that strategy
immediately follows from Corollary~\ref{corollary:exbetaeta} However, we omit such proofs since they can
can be formulated on untyped lambda terms and thus are orthogonal to
the specifics of dependently typed systems. Instead, we redirect
readers to \citet{factorization-essentially,
takahashi-parallel-reduction} for the details.

@@

>>>>

  The only interesting case is \Rref{I-Red}. Given that
  $[[A => B0]]$ and $[[Interp I i B0 S]]$, we need to show
  for all  $[[B1]]$ such that $[[A => B1]]$, we have $[[Interp I i B1
  S]]$. By the diamond property of parallel reduction
  (Lemma~\ref{lemma:pardiamond}), there exists some term $[[B]]$ such
  that $[[B0 => B]]$ and $[[B1 => B]]$. By the induction hypothesis,
  we deduce $[[Interp I i B S]]$ from $[[B0 => B]]$ and $[[Interp I i
  B0 S]]$. By \rref{I-Par} and $[[B1 => B]]$, we conclude that
  $[[Interp I i B S]]$.

@@

\section{Existence of $\beta$-normal forms}
\label{sec:extension}
In this section, we sketch out how the logical relation from
Section~\ref{sec:logreldep} can be extended to show the existence of
$\beta$ normal forms for (open and closed) well-typed terms.
\scw{Why is this a sketch? We've done the whole proof and we will
explain it all, right?}

The goal of this section is to prove that all well-typed terms have
$\beta$-normal forms. In other words, we can repeatedly use our parallel
reduction relation to reduce any term to a unique normal form, where no
further reductions can be applied. This result can be used to show that our
type conversion relation is decidable, as computation of the normal form will
always terminate.

\begin{figure}[h]
  \[
    \begin{array}{llcl}
       \beta\text{-}\mathit{neutral\ terms} &
      [[e]] & ::= & [[i]]\ |\ [[e f]]\ |\ [[J e f f f]]\ |\ [[if e f
                    f]] \\ \\
      \beta\text{-}\mathit{normal\ terms} &
      [[f]] & ::= & [[e]]\ |\ [[Set i]]\ |\ [[Void]]\ |\ [[Pi f f]]\
                    |\ [[f ~ f : f]]\\
            & & |   & [[\ f]]\ |\ [[refl]]\ |\ [[Bool]]\ |\ [[true]]\ |\ [[false]]
    \end{array}
  \]
  \caption{$\beta$-neutral and normal forms}
  \label{fig:nenf}
\end{figure}

The syntactic forms $[[e]]$ and $[[f]]$ (Figure~\ref{fig:nenf}) capture the
neutral terms and normal forms with respect to $\beta$-reduction. We sometimes
use the judgment forms $[[ne a]]$ and $[[nf a]]$ to indicate that there exists
$[[e]]$ or $[[f]]$ such that $[[a = e]]$ or $[[a = f]]$.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal form]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  natural numbers as term variables. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:crelrec} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemmas~\ref{lemma:crelrec} and \ref{lemma:cr2tyrec}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

Due to the non-deterministic nature of parallel reduction, we need to take a
few more steps to convert the existence of $\beta$-normal form into a decision
procedure for type conversion. More specifically, we can show that a
deterministic evaluation strategy such as leftmost-outermost reduction can
always find the $\beta$-normal form if there exists one. Then the termination
of that strategy immediately follows from Corollary~\ref{corollary:exbetanf}
However, we omit such proofs since they can can be formulated on untyped
lambda terms and thus are orthogonal to the specifics of dependently typed
systems. Instead, we redirect readers to \citet{factorization-essentially,
  takahashi-parallel-reduction} for the details.

@@

\section{Existence of $\beta\eta$-normal forms}
\scw{relocated text from previous section. Not a complete description yet.}

We first extend parallel reduction to include $\eta$ reduction for functions.
\begin{center}
  \drule[width=2.5in]{P-AbsEta}
\end{center}

\begin{corollary}[Existence of $\beta\eta$-normal form]
\label{corollary:exbetaeta}
If $[[G |- a : A]]$, then $[[a]]$ has $\beta\eta$-normal form.
\end{corollary}
\begin{proof}
  From Corollary~\ref{corollary:exbetanf}, we can easily conclude the
  existence of $\beta\eta$-normal forms for well-typed terms since
  $\eta$-reduction preserves $\beta$-normal form (implied by
  Lemma~\ref{lemma:parnenf}) and also strictly decreases the size of
  the term (evident from \rref{P-AbsEta}).
\end{proof}
